---
title: "MLB Analysis - Project Draft 1"
author: "Group 6: Nick Keohan, Harrison Flynn, and Shaun Harrington"
date: "March 28, 2021"
output:
  html_document: 
    fig_width: 8
    fig_height: 10
    code_folding: hide
---

</br>
</br>
</br>

Data from this analysis is thanks to Sean Lahman. The R package 'Lahman' is used to load this data. For more information see the documentation at  http://www.seanlahman.com/files/database/readme2017.txt.

</br>
</br>

The data will be formatted and visualized by means of the 'tidyverse' and 'ggplot2' R packages and Tableau.



```{r Setup, message=FALSE, warning=FALSE}

#Libraries
{
  library(Lahman)
  library(tidyverse)
  library(ggplot2)
  library(ggrepel)
  library(ggstar)
  library(directlabels)
  library(scales)
  library(gghighlight)
  library(hexbin)
}

```


### Top Team Managers

We'll take a look at the most successful team managers since 1884. This will be done by looking at most wins and most World Series Championships.

By manipulating the Managers and Teams tables in the 'Lahman' package, we can get our dataset started. 

```{r Mgr Data Prep, echo=TRUE, message=FALSE, warning=FALSE}

wins.byManager <- left_join(
      x = Managers,
      y = Teams,
      by = c("yearID", "lgID", "teamID"), suffix = c("", ".y")
    ) %>% 
      select(-contains(".y")) %>% 
      mutate(
        DivWin = ifelse(DivWin == "Y", 1, 0),
        WCWin = ifelse(WCWin == "Y", 1, 0),
        LgWin = ifelse(LgWin == "Y", 1, 0),
        WSWin = ifelse(WSWin == "Y", 1, 0)
      ) %>% 
      group_by(playerID, teamID) %>% 
      summarize(
        year.bgn = min(yearID),
        year.end = max(yearID),
        W = sum(W),
        L = sum(L),
        DivWin = sum(DivWin),
        WCWin = sum(WCWin),
        LgWin = sum(LgWin),
        WSWin = sum(WSWin)
      )

ws.winManager <- wins.byManager %>% 
  #group_by(playerID, teamID) %>% 
  group_by(playerID) %>% 
  summarize(
    year.bgn = min(year.bgn),
    year.end = max(year.end),
    W = sum(W),
    L = sum(L),
    DivWin = sum(DivWin),
    WCWin = sum(WCWin),
    LgWin = sum(LgWin),
    WSWin = sum(WSWin)
  ) %>% 
  filter(!is.na(WSWin), WSWin > 0) %>% 
  mutate(
    WSWin.Avg = WSWin / (year.end - year.bgn + 1)
  )

```

\newpage

Determining the performance of managers can be from numerous perspectives. We'll begin with the metric used to determine the best team each year: World Series Championships.

Below are the top 5 coaches by World Series Championships. Showing the cumulative championships by manager, irrespective of the team that they managed.

```{r Mgr - WS, echo=TRUE, fig.height=9, fig.width=8, message=FALSE, warning=FALSE}


#Time Series Chart
{
  temp <- left_join(
    x = Managers,
    y = Teams,
    by = c("yearID", "lgID", "teamID"), suffix = c("", ".y")
  ) %>% 
    select(-contains(".y")) %>% 
    mutate(
      DivWin = ifelse(DivWin == "Y", 1, 0),
      WCWin = ifelse(WCWin == "Y", 1, 0),
      LgWin = ifelse(LgWin == "Y", 1, 0),
      WSWin = ifelse(WSWin == "Y", 1, 0)
    ) %>% 
    filter(playerID %in% ws.winManager$playerID) %>% 
    group_by(playerID) %>% 
    arrange(yearID, teamID) %>% 
    add_count() %>% 
    mutate(
      Total.WSWin = sum(WSWin),
      Running.WSWin = cumsum(WSWin),
      WSWin.Percent = Total.WSWin / n,
      team.chg = ifelse(teamID == lag(teamID), 0, 1),
      Win.Percent = W / G * 100
    ) %>% #select(yearID, playerID, teamID, team.chg)
    replace_na(list(team.chg = 1)) %>% 
    filter(
      playerID %in% (ungroup(.) %>% 
                       select(playerID, Total.WSWin) %>%
                       distinct() %>% 
                       top_n(5, Total.WSWin))$playerID
    ) %>% 
    left_join(
      x = ., y = People, by = "playerID", suffix = c("", ".y")
    ) %>% 
    left_join(
      x = ., y = Teams, by = "teamID", suffix = c("", ".y")
    ) %>% 
    select(yearID, playerID, teamID, W, WSWin, WSWin.Percent, Win.Percent,
           Running.WSWin, Total.WSWin, team.chg, nameFirst,
           nameLast, name) %>% 
    distinct %>% 
    mutate(
      team.chg = ifelse(team.chg == 1, TRUE, FALSE),
      nameFull = paste(nameFirst, nameLast, sep = " ")
    )
  
  temp %>% 
    ggplot(aes(x = yearID, y = Running.WSWin, color = nameFull)) +
    geom_step(direction = "hv", size = 1) +
    geom_star(aes(starshape = as.factor(team.chg), fill = nameFull,
                  size = Win.Percent),
              color = "black") +
    geom_label_repel(aes(label = ifelse(team.chg == 1, name,""),
                         fill = nameFull),
                    nudge_y = .25, size = 2.5, color = "black",
                    force = .1, max.overlaps = 15) +
    geom_dl(aes(label = paste(nameFirst, nameLast, sep = "\n")), 
            method = list(#dl.trans(x = x + 0), 
                          dl.trans(y = y + .25),
                          "smart.grid", rot = 0, cex = .9),
            size = 2) +
    scale_size(range = c(2,5)) +
    scale_starshape_manual(values = c(28, 29)) +
    scale_y_continuous(breaks = c(0:7)) +
    scale_x_continuous(n.breaks = 10) +
    scale_fill_discrete(guide = FALSE) +
    ggtitle(
      "Top 5 Team Managers by World Series Championships",
      "1884 - 2019"
    ) +
    xlab("Year") +
    ylab("Total World Series Championships") +
    labs(
      starshape = "First Year at Team",
      size = "Season Winning Percentage",
      color = "Team Manager"
    ) +
    theme_bw() +
    theme(
      plot.title = element_text(hjust = 0.5),
      plot.subtitle = element_text(hjust = 0.5),
      legend.key.size = unit(2,"line")
    ) +
    guides(
      size = guide_legend(override.aes = list(starshape = 28))
    )
  
  
}


```

</br>

\newpage

Another perspective is to consider the total number of wins a manager is able to lead their teams to.

</br>

Here is the code to prep this dataset:

```{r Mgr Total Wins Prep, message=FALSE, warning=FALSE}




total.wins <- left_join(
  x = Managers,
  y = Teams,
  by = c("yearID", "lgID", "teamID"), suffix = c("", ".y")
) %>% 
  select(-contains(".y")) %>% 
  mutate(
    DivWin = ifelse(DivWin == "Y", 1, 0),
    WCWin = ifelse(WCWin == "Y", 1, 0),
    LgWin = ifelse(LgWin == "Y", 1, 0),
    WSWin = ifelse(WSWin == "Y", 1, 0)
  ) %>% 
  # filter(playerID %in% (wins.byManager %>%
  #                         ungroup %>% 
  #                         top_n(10, W))$playerID) %>% 
  group_by(playerID) %>% 
  arrange(yearID, teamID) %>% 
  add_count() %>% 
  mutate(
    Total.Win = sum(W),
    Running.Win = cumsum(W),
    team.chg = ifelse(teamID == lag(teamID), 0, 1),
    Win.Percent = W / G * 100
  ) %>% #select(yearID, playerID, teamID, team.chg)
  replace_na(list(
    team.chg = 1, DivWin = 0, WCWin = 0, LgWin = 0, WSWin = 0
  )) %>% 
  left_join(
    x = ., y = People, by = "playerID", suffix = c("", ".y")
  ) %>% 
  left_join(
    x = ., y = Teams, by = "teamID", suffix = c("", ".y")
  ) %>% 
  select(yearID, playerID, teamID, lgID, W, Win.Percent, DivWin, LgWin, WSWin,
         Running.Win, Total.Win, team.chg, nameFirst,
         nameLast, name) %>% 
  distinct %>% 
  mutate(
    team.chg = ifelse(team.chg == 1, TRUE, FALSE),
    nameFull = paste(nameFirst, nameLast, sep = " ")
  )

total.wins.top <- total.wins %>%
  filter(playerID %in% (wins.byManager %>%
                          ungroup %>%
                          top_n(10, W))$playerID)




        
      
```




```{r Mgr Total Wins Chart, fig.height=9, fig.width=8, message=FALSE, warning=FALSE}

total.wins %>% 
  mutate(WSWin = ifelse(WSWin == 1, TRUE, FALSE)) %>% 
  ggplot() +
  geom_line(
    aes(x = yearID, y = Running.Win, color = nameFull)
  ) +
  gghighlight(
    #aes(x = yearID, y = Running.Win, color = nameFull),
    max(Running.Win), max_highlight = 10,
    use_direct_label = FALSE
  ) +
  #facet_wrap(. ~ lgID) +
  geom_star(
    aes(
      x = yearID, y = Running.Win, starshape = as.factor(WSWin),
      fill = nameFull, size = as.factor(WSWin)
    ),
    color = "black"
  ) +
  geom_text_repel(
    aes(
      x = yearID, y = Running.Win, label = ifelse(WSWin == 1, name,"")
    ),
    #nudge_y = .25,
    size = 2, 
    color = "gray15"
    #force = .1, max.overlaps = 15,
    #direction = "x"
  ) +
  geom_dl(
    aes(
      x = yearID, y = Running.Win, color = nameFull,
      label = paste(nameFirst, nameLast, sep = " ")
    ), 
    method = list(
      dl.trans(x = x - .05), 
      dl.trans(y = y + .15),
      "last.points", rot = 65, cex = .75
    ),
    size = .75
  ) +
  #scale_size(range = c(0,2), guide = FALSE) +
  scale_starshape_manual(
    values = c(15, 1), guide = FALSE
  ) +
  scale_y_continuous(
    n.breaks = 10,
    limits = c(0, max(total.wins.top$Running.Win)+300),
    labels = scales::comma
  ) +
  scale_x_continuous(
    n.breaks = 10,
    limits = c( min(total.wins.top$yearID), max(total.wins.top$yearID)+1)
  ) +
  scale_fill_discrete(guide = FALSE) +
  scale_size_manual(
    values = c(0,3), guide = FALSE
  ) +
  ggtitle(
    "Top 10 Team Managers by Total Career Wins",
    "1884 - 2019"
  ) +
  xlab("Year") +
  ylab("Total Cumulative Wins") +
  labs(
    size = "World Series Champion",
    color = "Team Manager"
  ) +
  theme_bw() +
  theme(
    plot.title = element_text(hjust = 0.5),
    plot.subtitle = element_text(hjust = 0.5),
    legend.key.size = unit(2,"line")
  ) +
  guides(
    size = guide_legend(override.aes = list(starshape = 1))
  )

```

\newpage 

### Regular vs. Post Season Performance 

```{r Post Seas Chart, message=FALSE, warning=FALSE, fig.height=9}


    Teams %>% 
      mutate(
        teamYear = paste(teamID, yearID, sep = "\n"),
        Win.Percent = W / G * 100,
        DivWin = ifelse(DivWin == "Y", "Division Winner", "Division Loser"),
        LgWin = ifelse(LgWin == "Y", "League Winner", "League Loser"),
        WSWin = ifelse(WSWin == "Y", "World Series Champion", "Non-Champion")
      ) %>% 
      filter(!is.na(DivWin), !is.na(LgWin), !is.na(WSWin)) %>% 
      #ggplot(aes(x = L, y = W, color = as.factor(Rank))) +
      ggplot(aes(x = L, y = W, fill = WSWin, starshape = WSWin)) +
      geom_star(size = 3) +
      facet_grid(
        #WSWin ~ DivWin + LgWin, scales = "fixed", margins = FALSE
        LgWin ~ DivWin, scales = "fixed", margins = FALSE
      ) +
      
      #50% Win Percentage
      geom_segment( 
        aes(x = 40, xend = 100, y = 40, yend = 100),
        inherit.aes = FALSE, linetype = "dotted"
      ) + 
      geom_text(
        aes(x = 100, y = 98, label = "50% Winning Percentage"),
        inherit.aes = FALSE, angle = 43.5,
        color = "gray40", size = 3
      ) +
      
      #110 Number of Games
      geom_segment( 
        aes(x = 70, xend = 45, y = 40, yend = 65),
        inherit.aes = FALSE, linetype = "dashed"
      ) + 
      geom_text(
        aes(x = 70, y = 43, label = "110 Games Played"),
        inherit.aes = FALSE, angle = -43.5,
        color = "gray40", size = 3
      ) +
      
      #160 Number of Games
      geom_segment( 
        aes(x = 105, xend = 50, y = 55, yend = 110),
        inherit.aes = FALSE, linetype = "dashed"
      ) + 
      geom_text(
        aes(x = 59, y = 108, label = "160 Games Played"),
        inherit.aes = FALSE, angle = -43.5,
        color = "gray40", size = 3
      ) +
      scale_x_continuous(n.breaks = 4) +
      scale_y_continuous(n.breaks = 6) +
      ggtitle(
        "Post-Season Performance Against Total Record",
        "1969 - 2019"
      ) +
      xlab("Total Season Losses") +
      ylab("Total Season Wins") +
      labs(
        fill = "World Series Champion",
        starshape = "World Series Champion"
      ) +
      theme_bw() +
      theme(
        plot.title = element_text(hjust = 0.5),
        plot.subtitle = element_text(hjust = 0.5),
        legend.position = "bottom"
      ) +
      # scale_fill_gradient(
      #   low = "skyblue1", high = "midnightblue"
      # ) +
      scale_fill_manual(values = c("lightsteelblue3", "gold3")) +
      scale_starshape_manual(values = c(15, 1))
  

```

